---
title: "**Prediction of length-of-stay in an ICU bed**"
output: 
  html_document:
    css: css/main.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

# Useful packages
library(jsonlite)
library(tidyverse)
library(here)
library(survival)
library(lubridate)
library(MASS)
```


_Note: for now I used 100% random information of length-of-stay_

# Background
The <span style="color:#46c3a4">_**GreenBed**_</span> platform is supposed to provide information on the availability of ICU beds in Switzerland. At admission of a patient to the ICU, a member of the hospital staff will block the bed, the patient is put in. Further some baseline patient characteristics are provided to the platform. In a next step, using these patient characteristics together with some bed-level information and hospital-level data the platform predicts the length-of-stay of the patient and therefore a time window for which the bed will be occupied. This data will help with hospital resource management. As this platform can be accessed Swiss-wide and the data is anonymised, it further provides real-time information on availability of beds, which can help to reallocate patients inbetween cities or even cantons.

# Data 
The data used for the prediction of the length-of-stay will be collected by the platform itself. The usage of historical data for the initial predictions can be considered, so that predictions can be furnished right from the start. Note that the data recorded by the app will be admission-specific, and for the prediction we are interested in the bed itself, and not in the individual patient. This ensures that we are not working with actual sensitiv patient-data. The data collected by the platform will be the following:

- unique identifier of the bed (`ID_BED`).
- unique identifier of the hospital (`ID_HOSPITAL`)
- date of admission of the patient (`ADMISSION_DATE`)
- gender of the patent (`GENDER`)
- age of the patient (`AGE`) (_can be changed to birthdate instead, however due to data-privacy issue we prefere not to_)

```{r import-and-clean-up-data}
patients <- read.csv(file = "data/ICU_SWITZERLAND_MOCK_V1.csv")

n_sample <- nrow(patients)

set.seed(123)
# let's assume a patient is in an ICU bed an average of 7 days
avg_nb_days <- 7
randomly_allocate_a_length_of_stay <- rpois(n = n_sample,
                                            lambda = avg_nb_days)

patients <- cbind(randomly_allocate_a_length_of_stay, 
                  patients)
```

## Mock data structure explained

Python was used to build a synthetic data set. The sample size is `r prettyNum(n_sample, big.mark="'")`. 

## Predictors

After a literature review the following potential predictors were retrieved:

- tba 
- tba


# Training and Testing

To evaluate the models and in the long-term maybe send alerts when the model is not performing anymore, the data is split into a train and a testing sample. The training data is composed of 75% of the sample. Note that in the long term this should be changed to cross-validation (*e.g* splitting the data several times).

```{r, echo = TRUE}
# Split into train and test
set.seed(1)
train_rows <- sample(1:n_sample, size = .75*n_sample)
train_patients <- patients[train_rows, ]
test_patients <- patients[-train_rows, ]
n_train <- nrow(train_patients)
n_test <- nrow(test_patients)
```


# Description of potential models
Once the product is running, data is collected and lessons have been learned, more sophisticated models can be looked at ($e.g$ machine learning). For the time-being and for a first prototype simple parametric regression models should be used.

## Simple poisson regression model 
Since we want to predict the number of days a patient stays in the bed, this can be resolved with a simple count-data problem.
```{r poisson-model-training, echo = TRUE}
poisson_regression <- 
  glm(randomly_allocate_a_length_of_stay ~ Gender + Age + Ort + 
           zertifizierte..Betten,
      family = "poisson",data = train_patients)
```


## Negative-binomial regression model
An alternative to poisson regression are negative-binomial models.

```{r negbin-model-training, echo = TRUE, warning=FALSE}
negbin_regression <- 
  glm.nb(randomly_allocate_a_length_of_stay ~ Gender + Age + Ort + 
           zertifizierte..Betten,
          data = train_patients)
```

## Tree-based solutions

More non-parametric models should be looked at, once more data has been collected.

# Model performance and variable selection on the mock-data

To compare the model performances, let us use the root-mean-square error (RMSE). The RMSE is commonly used to measure differences between values predicted by a model and the values observed, where a lower score implies better accuracy. 

This model validation can also be used to update the models, add or drop variables, by judging whether they have had any predictive power.

Let's see how well that first poisson model performs:
```{r poisson-model-testing, echo = TRUE}
predict_los <- predict(poisson_regression, newdata = test_patients,
                       type = "response")
observed_los <- test_patients$randomly_allocate_a_length_of_stay

RMSE <- sqrt(sum((observed_los - predict_los)**2)/n_test)
RMSE
```


# Data-privacy issues to consider

To ensure the data being pseudo-anonymised, meaning making it impossible, or very hard to identify patients, we can consider the following further steps:

- Age-groups instead of exact age (<18, [18-29], [30-49], [50-64], [65-74], 75+)