---
title: "**Prediction of occupancy of an ICU bed**"
output: 
  html_document:
    code_folding: show
    css: css/main.css
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

# Useful packages
library(jsonlite)
library(tidyverse)
library(here)
library(survival)
library(lubridate)
library(MASS)
```

<img src = "logo.png" alt="My Logo" style="float:right;width:150px;height:150;">

# Background

The <span style="color:#2B7ACA">_**BlueBeds**_</span> platform is supposed to provide information on the availability of ICU beds in Switzerland. At admission of a patient to the ICU, a member of the hospital staff will block the bed, the patient is put in. Further some baseline patient characteristics are provided to the platform. In a next step, using these patient characteristics together with some bed-level information and hospital-level data the platform predicts the length-of-stay of the patient and therefore a time window for which the bed will be occupied. This data will help with hospital resource management. As this platform can be accessed Swiss-wide and the data is anonymised, it further provides real-time information on availability of beds, which can help to reallocate patients in between cities or even cantons.

# Data collection
The data used for the prediction of the length-of-stay (of the patient in the bed) will be collected by the platform itself. The usage of historical data for the initial predictions can be considered, so that predictions can be furnished right from the start. Note that the data recorded by the platform will be admission-specific, and for the prediction we are interested in the bed itself, and not in the individual patient. This ensures that we are not working with actual sensitive patient-data. The data collected by the platform will be the following (the selection is based on a literature review):

- unique identifier of the bed (`ID_BED`).
- unique identifier of the hospital (`ID_HOSPITAL`)
- date of admission of the patient (`ADMISSION_DATE`)
- date of discharge of the patient (`DISCHARGE_DATE`)
- length-of-stay of the patient (`LENGTH_OF_STAY`)
- gender of the patient (`GENDER`)
- age of the patient (`AGE`) (_can be changed to birth date instead, however due to data-privacy issue we prefer not to_)
- co-morbidities or preconditions of the patient (`COMORB`). This can be a text input, with listed co-morbidities separated by commas, so that the model can either use a binary variable for presence of preconditions (`PRESENCE_COMORB`) or even the number of different preconditions (`NB_COMORB`). 
- some Covid-19 related severity score as for example the [Brescia-COVID Respiratory Severity Scale](https://www.mdcalc.com/brescia-covid-respiratory-severity-scale-bcrss-algorithm)
- Covid-19 related symptoms or simply again the number of symptoms (`NB_SYMPTOMS`)


The hospital staff (doctor or nurse) interacts with the platform only twice: 

1. To find a bed and admit the patient to this bed. At this stage the doctor or nurse, sets the bed to occupied, fills in the baseline characteristics of the patient. The demographics on the bed, hospital, etc are automatically recorded by the platform. Also the admission date is automatically filled in by the platform.

2. When the patient is discharged from the bed. At this second stage, the doctor or nurse sets the bed to free and fills in the final stage of the patient (dead, alive, other).


```{r import-and-clean-up-data, warning= FALSE, message = FALSE}
patients <- readxl::read_xlsx(path = "data/TRAINING_DURATION_OF_STAY.xlsx") %>% 
  rename(PATIENT_ID = "...1",
         LENGTH_OF_STAY = "DUATION OF STAY",
         NB_COMORB = "# of pre_conditions",
         NB_SYMPTOMS = "# of symptoms",
         AGE = Age, 
         GENDER = Gender)

n_sample <- nrow(patients)

# set.seed(123)
# patients <- cbind(randomly_allocate_a_length_of_stay, 
#                   patients)
```

# Training and testing of the model

To evaluate the model and, in the long-term, send alerts if the model is under-performing, it is important to schedule reguare tests. Everytime a new patient is discharged from a bed, the complete record of the ICU bed is added to the data base. Next, the prediction model is updated using the new data point(s). Is is extremely important to alwayse split the available data into training and testing data; train the model on one part and test it on the remaining. Different approaches can be followed here ($e.g.$ K-fold cross-validation or bootstrap cross-validation).

# Description of potential models
Once the platform is running, data is collected and lessons have been learned, more sophisticated models can be looked at ($e.g$ machine learning). For the time-being and for a first prototype simple parametric regression models should be sufficient

## Simple Poisson regression model 
Since we want to predict the number of days a patient stays in the bed, this can be resolved with a simple count-data problem. The most-commonly used method for such data is the [Poisson regression](https://en.wikipedia.org/wiki/Poisson_regression).

## Negative-binomial regression model
An alternative to Poisson regression are [negative-binomial models](https://en.wikipedia.org/wiki/Negative_binomial_distribution). These are generalisation of the Poisson model, loosening some of the underlying assumptions of the latter. 

## Tree-based solutions
Non-parametric models can be looked at, once more data has been collected; for example models based on descision trees or machine learning in general.

## Further modelling strategies
An add-on output of the platform can be the prediction of the not only the time the patient leaves the bed, but further the chances of leaving the bed alive. This can be done using more sophisticated _time-to-event models_ after the data on the end-state of the patients at the time of discharge of the ICU bed is retrieved. As such, the platform can be seen as additional source for robust data collection to better understand the course of the disease. 

# Model building, performance testing and variable selection on the mock-data
To better illustrate the usage of the prediction model, a mock data set has been simulated, which can be downloaded [here](https://github.com/rachelHey/versus-virus-hack-length-of-stay-prediction/blob/master/data/TRAINING_DURATION_OF_STAY.xlsx). The data has `r prettyNum(n_sample, big.mark="'")` rows with information on hospital admissions. To build the data, and make it as realistic as possible, several sources with information on Covid-19 patients were used. For the clinical characteristics, their symtoms, etc, this [JAMA paper](https://jamanetwork.com/journals/jama/fullarticle/2761044) was a good source, especially [Table 1](ttps://cdn.jamanetwork.com/ama/content_public/journal/jama/938419/joi200019t1.png?Expires=2147483647&Signature=qAtakdiLVahD-1x8zl9gyZ07WD6M2lmSQPs3vG60dc6K4y5cxTDa7DbNrqO1biI2uiwC0fnIOrfzLhh4P5q3-IYGOi5jFp5Scj6PYS9xYiFc-HHDuWeq~imiMQnhGhKytxqPRYwat9uywS43a7SbSSj88DJgSNZcp2yq5GePiC6HVPAImWsVtSg5X7xNZirkp9gstf4doM54DkODLhAz86gjCInulisjJyYhFZxEv~6bvxYFxH3ARUPCktCA~voJHwD~jFbNfJLymecdjNFjzzRQqUs25~OnNuxq7xMm3FPuz3bqJG~1Z8-9K1QgomEx6EeEJHmV9kDNI3Dv8nRqDw__&Key-Pair-Id=APKAIE5G5CRDK6RD3PGA).

The illustrative model will include the following variables: the age of the patient, their gender, the number of Covid-19 related symptoms and the number of comorbidities.

## Model building
To build the prediction model, 75% of the data is used as training sample, as is illustrated in the following `R`-code chunk:
```{r, echo = TRUE}
# Split into train and test
set.seed(1)
train_rows <- sample(1:n_sample, size = .75*n_sample)
train_patients <- patients[train_rows, ]
test_patients <- patients[-train_rows, ]
n_train <- nrow(train_patients)
n_test <- nrow(test_patients)
```

Then, the actual poisson regression model is fitted on this training data:
```{r poisson-model-training, echo = TRUE}
poisson_regression <- 
  glm(LENGTH_OF_STAY ~ GENDER + AGE + NB_SYMPTOMS + NB_COMORB,
      family = "poisson",data = train_patients)
```

Alternatively, a negative binomial regression model is fitted: 
```{r negbin-model-training, echo = TRUE, warning=FALSE}
negbin_regression <- 
  glm.nb(LENGTH_OF_STAY ~ GENDER + AGE + NB_SYMPTOMS + NB_COMORB,
         data = train_patients)
```


## Model testing

To compare the model performances, we will use the [root-mean-square error](https://en.wikipedia.org/wiki/Root-mean-square_deviation) (RMSE), for symplicity. The RMSE is commonly used to measure differences between values predicted by a model and the values observed, where a lower score implies better accuracy. 

Let's see how well that first poisson model performs:
```{r poisson-model-testing, echo = TRUE}
predict_los <- predict(poisson_regression, newdata = test_patients,
                       type = "response")
observed_los <- test_patients$LENGTH_OF_STAY

RMSE <- sqrt(sum((observed_los - predict_los)**2)/n_test)
```

For this first model, RMSE = `r round(RMSE, 2)`.

This model validation step can also be used to update the models, add or drop variables, by judging whether they have had any predictive power.

## Prediction of occupancy length for imaginary patients
The following Table shows the basic characteristics of ten imaginary patients. The predictions given by the model on the number of days until the bed will be available, are found in the last column. 
```{r}
tab <- data.frame(GENDER = c("Female", "Male", "Male", "Male", "Female",
                             "Male", "Female", "Female", "Female",  "Male"),
                  AGE = c(49, 84, 76, 25, 71, 66, 34, 80, 91, 62),
                  NB_SYMPTOMS = c(1, 6, 2, 4, 5, 5, 1, 3, 3, 6),
                  NB_COMORB = c(1, 1, 1, 3, 0, 1, 2, 1, 1, 2))

prediction_imaginary <- predict(poisson_regression, newdata = tab, 
                                type = "response") %>% 
  round(2)

tab <- cbind(tab, prediction_imaginary)
colnames(tab)[5] <- "Prediction length-of-stay"
kableExtra::kable(tab, 
                  caption = "Ten imaginary patients, together with a prediction for their length-of-stay.") %>% 
  kableExtra::kable_styling()
```


# Data-privacy issues to consider

To ensure the data being pseudo-anonymised, meaning making it impossible, or very hard to identify patients, the following steps have to be considered:

- The data stored by the platform and used in the model are ICU bed-specific and do not store sensible patient data. 
- Age-groups instead of exact age (<18, [18-29], [30-49], [50-64], [65-74], 75+).